Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-26 icfp.tex
pass 5

{\small
  \input{prototype/core-plugin-results-2}
  }
%
The results indicate that our mostly direct implementation of Linear Core is
successful in accepting the vast majority of the thousands of intermediate
programs produced by GHC when compiling libraries that make extensive use of
linear types.
%
However, Linear Core does not accept every single program produced by the
optimisation pipeline of GHC. Our manual analysis of rejected programs revealed
excellent results. Programs considered ill-typed by the Linear Core
implementation include:
%
\begin{itemize}
    \item Programs we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
        Namely, programs which scrutinize a variable, but then use the variable in the
        case alternatives of the case scrutinizing it. As discussed in
        Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
        understood as linear as long as applications of linear functions binding these
        variables are reduced \emph{call-by-need}, which is not the case in our system.

@@

    \item Programs which \emph{actually violate linearity}, in that the term
        violates the linear type it is annotated with. These are the programs
        we ultimately want to be able to reject by having a linearly typed Core
        validate intermediate programs. We found both a program which outright
        discarded a linear resource, and that rewrite rules implementing stream
        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199}
        (among others), could produce linearly-invalid programs (in one such case,
        the type of \lstinline{build} was not general enough to accommodate linearity).
\end{itemize}
%
Overall, the programs rejected by the Linear Core plugin,
%, in \texttt{linear-base},
besides validating that our implementation is faithful to the
$\lambda^\pi_\Delta$ system of insofar as programs that should not be accepted
are deemed ill-typed, provide further insight into the remaining
details required to fully typecheck linearity in a mature optimising compiler.

>>>>

{\small
  \input{prototype/core-plugin-results-2}
  }
%
The results indicate that our mostly direct implementation of Linear Core is
successful in accepting the vast majority of the thousands of intermediate
programs produced by GHC when compiling libraries that make extensive use of
linear types.
%
However, Linear Core does not accept every single program produced by the
optimisation pipeline of GHC. Our manual analysis of rejected programs revealed
excellent results. Programs considered ill-typed by the Linear Core
implementation include:
%
\begin{itemize}
    \item Programs we expected GHC to produce, but know are not seen as linear by $\lambda^\pi_\Delta$.
        Namely, programs resulting from the reverse binder swap which
        scrutinize a variable and then use it again in the case alternatives.
        As discussed in Section~\ref{sec:reverse-binder-swap-considered-harmful}, these
        programs could be understood as linear as long as all applications of
        linear functions are reduced \emph{call-by-need}, but our system
        conservatively rejects it in light of the flexible evaluation strategy
        used by optimizing compilers.

@@

    \item Programs which actually violate linearity, in that the term
        violates the linear type it is annotated with. These are the programs
        we ultimately want to spot and reject by having a linearly typed Core
        validate intermediate programs. We found both a program which outright
        discarded a linear resource and that rewrite rules implementing stream
        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199}
        (among other rules) could produce linearly-invalid programs (in one such case,
        the type of \lstinline{build} was not general enough to accommodate linearity).
\end{itemize}
%
Overall, the programs rejected by the Linear Core plugin,
%, in \texttt{linear-base},
besides validating that our implementation is faithful to the
$\lambda^\pi_\Delta$ system insofar as programs that should not be accepted
are deemed ill-typed, provide further insight into the remaining
details required to fully typecheck linearity in a mature optimising compiler.
